--- 
title: "Electoral Margins"
author: "Andus Kong, Jiyeon Chang"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
---

## Background

Against the backdrop of the upcoming election, the media has given a lot of attention to races that are expected to be particularly competitive. Given the large share of international students in the class, we thought it would be interesting to provide some background as to what the electoral map for the presidential election looks like, and how margins have worked out in past elections.

```{r, include=FALSE}
setwd("/Users/siyeona/Downloads/mphil_data")
#install.packages("mapproj")
library(dplyr)
library(tidyverse)
library(haven)
library(car)
library(ggplot2)
library(maps)
library(mapproj)
library(cowplot)
library(readxl)
library(magrittr)
```

The data for the analysis is taken from the MIT Election Lab (https://electionlab.mit.edu/) which provide information on the winners and the votes they received, as well as information on turnout. For the purpose of this analysis I look at data from 2000 to 2016.

```{r, include=FALSE}
load(file = "1976-2016-president.RData"); elect_pres<-x; rm(x) # presidential
turnout<-read_excel("turnout.xlsx")
turnout<-turnout %>%
  select(c("year","state","vep_highest_office")) %>%
  filter(state!="United States")

elect_data<-elect_pres %>%
  select(c("year","state","state_po","candidate","party","totalvotes","candidatevotes"))

pres_top2<- elect_data %>%
  group_by(state,year) %>%
  arrange(desc(candidatevotes),.by_group = TRUE) %>%
  filter(row_number() %in% c(1,2)) %>%
  mutate(share = candidatevotes/totalvotes) %>%
  mutate(margin = share-dplyr::lead(share))

pres_winner<- pres_top2 %>%
  group_by(state,year) %>%
  filter(row_number() %in% c(1)) %>%
  mutate(margin_cont = ifelse(party=="republican",margin,-(margin)))

margins<-merge(pres_winner, turnout, by=c("state","year"))

```

The maps packages does BLAH BLAH (ADD). The basic 
```{r}
us_states <- map_data("state")
head(us_states)
margins$state <- tolower(margins$state)
names(us_states)[names(us_states) == "region"] <-"state"

df.margin <- merge(margins, us_states, sort = FALSE, by = "state")
```

```{r}
margin16 <- df.margin %>%
  filter(year==2016)
```

As a starter, let's look at how states voted in the 2016 elections. By specifying scale_fill_manual, and assigning the colors conventionally applied to the two main political parties in the United States, we get a map of the electoral outcome.
```{r}
# change legend aesthetics.. should be much smaller and blend in with the map
# center title.
p_party <- ggplot(data = margin16,
            aes(x = long, y = lat,
                group = state, fill = party))
col_party <- c("blue", "red")

p1<-p_party + geom_polygon(color = "gray90", size = 0.1) + 
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  # guides(fill = FALSE) +
  scale_fill_manual(values=col_party)+
  labs(title="Election outcome 2016") +
  theme_map()

p1
```

The above map gives us an idea of who carried the state, but not by what margin. To do this, we modify the fill parameter from party to margin, which is a continuous variable.
```{r}
# change it so that depending on the winning party, the margin should be dark blue if won by a big margin by democrat and dark red vice versa. Sharp margin should be light. <-- for very light areas, overlay the actual values.

p_margin <- ggplot(data = margin16,
            aes(x = long, y = lat,
                group = state, fill = margin))

p_margin + geom_polygon(color = "gray90", size = 0.1) + 
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  # guides(fill = FALSE) +
  labs(title="margin by state 2016",fill="percent") +
  scale_fill_gradient(low="red",high="white")+
  theme_map()

```
@ANDUS: any ideas for different color ??

This map gives us a better of idea where the race was particularly close, but given that the range of margins is quite large (from min X to max Y), it is not as useful in distinguishing, say, races that had a margin < 3%.

For now, let's break down the margin by who won..

```{r}
# same map as above but distinguishes between the margin when it leans democrat vs. republican. I.e. margins are red if republicans won, and blue if democrats won.
p_margin <- ggplot(data = margin16,
            aes(x = long, y = lat,
                group = state, fill = margin_cont))

p_margin + geom_polygon(color = "gray90", size = 0.1) + 
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  # guides(fill = FALSE) +
  labs(title="margin by state 2016",fill="percent") +
  scale_fill_distiller(palette = "RdBu")+
 # scale_fill_gradient(low="red",high="white")+
  theme_map()
```

ADD: a map where you only look at places with margin <3% and plot the actual values of margins on the map
```{r}
```


ADD
```{r}
#parallel coordinate plots with states on the y axis and margin at each year on the x axis as an interactive plot

```
ADD
```{r}
# cleveland dot plot of margins in 2016, with overlaid with turnout in 2016.
```

